% Copyright 2024- Jay Jay Billings. Some rights reserved.
\documentclass{article}
\usepackage{graphicx} % Required for inserting images
\usepackage{hyperref}
\usepackage{tabularx}

\title{For sale: A beautiful forest!}
\author{Jay Jay Billings, Ph.D.}
\date{\today}

\begin{document}

\maketitle

\textbf{This is the stub of an article that never happened. A week after throwing down these thoughts, we decided to keep the forest.}\newpage

Those who follow my annual updates (link: annual update) may recall that we were originally planning to build a house when we moved to Northern Virginia (NoVA). We love forests, so we bought a heavenly 37.78-acre spot north of Hillsboro in 2022. We ultimately decided to buy instead of build, and now we're hoping to sell our forest. In this post, I'll tell you all about it! 

\section*{The Forest}

\section*{History}

\section*{Conservation}

\section*{Forest Farming and Glamping}

\section*{Buy vs Build}

There's not a lot to this, really. We decided to buy instead of build because it worked better with our budget and timeline. We needed to have a home in NoVA by August 2023, which was based on when my oldest daughter started kindergarten. As we talked to more builders, we burnt up resources and had to face the reality that building a custom home was time-consuming and expensive. I was not willing to sacrifice our timeline because it meant impacting my daughter's education. It started making a lot of sense to stop trying to build and instead find a nice house. 

And finding a nice house in Loudoun County is pretty easy! Buying our current home only took eight weeks working with a great husband and wife realtor team who support one of the builders we met. It was a fraction of the cost, and it is very similar to what we would have built anyway. We closed in May 2023, three months ahead of when we wanted to be in NoVA, and I had a place to stay just in time for the return to office mandate at Amazon. It was extremely convenient, even though we'll have to plant a new forest on this property, and we have some customization to do. 

\section*{Realtor Info}

Drew and Chad - Whitetail properties

\end{document}
