% Copyright 2023 Jay Jay Billings. Some rights reserved.
\documentclass{article}
\usepackage{graphicx} % Required for inserting images
\usepackage{hyperref}
\usepackage{tabularx}

\title{Outliving your mind, part 2: Saltines}
\author{Jay Jay Billings, Ph.D.}
\date{\today}

\begin{document}

\maketitle

\textit{In my \href{https://jayjaybillings.com/2023/07/06/outliving-your-mind-part-1-the-secret/}{last post} on my father's battle with dementia, I shared how he kept it secret for nearly twenty years. In this post, I'll discuss how his symptoms progressed.}

There are many excellent websites that offer detailed descriptions of symptoms and timelines for various conditions and diseases that cause cognitive decline and impairment. However, I often find it difficult to correlate that content to what I witnessed with Dad or other friends and family. Like in the case of a family friend who shoved his caregivers aside, jumped on a lawnmower, flipped the bird, and yelled ``You'll never catch me, mother@\%\&\#\^\@*s!!!'' as he rode into the sunset. Which set of symptoms on \href{https://www.mayoclinic.org/diseases-conditions/dementia/symptoms-causes/syc-20352013}{the Mayo Clinic dementia page} covered that behavior?

I offer the narrative below as a sort of case study of the way cognitive impairment progresses, or at least how it happened for my father, and how dementia can manifest in an individual. I found the quick evolution of Dad's condition during the last three years of his life to be especially important because it was the most challenging, confusing, and undocumented phase. I'll offer a warning that, upon review, I found some parts of the narrative tough to read because this is ultimately a sad and confusing matter. 

\section*{Crackers}

I wondered if the scene in front of me was a sign that Dad had taken up a form of ``single septuagenarian grocery art.'' It was pretty in a way - colorful, possessing no specific pattern in its assembly, and imposing. Arranged before me from floor to ceiling and from the corner of the hall clear into the kitchen was a 7-foot tall by 10-foot long wall of empty and neatly arranged saltine boxes blocking his back door.

It was 2008 or so. I was visiting Dad after a long break. We had argued the previous Christmas because I wanted to keep my trip short. When I eventually made it home to see the collage of Premium, Zesta, and Food Club cracker boxes, I wasn't surprised by the stack. He had already started the project the last time I visited because his back door had a broken lock. He put a few boxes in front of the door in case someone broke into his trailer. It wasn't the worst idea if you needed something simple until your lock was fixed: an intruder pops open the back door, knocks the boxes out of the way as they do, can't put them back like they originally were, and thus leaves you some evidence that someone came through the broken door. Of course, I offered to fix the lock, but Dad assured me that he had a guy who was going to fix it soon. So, I was aware of the boxes but surprised by the enormity of the installation.

Dad lived in that trailer for eight years. He never fixed the lock. Every year, the people who were going to rob him increased in number and ferocity, at least as he told it, but no one ever broke in or even opened the door.

When he moved in 2016, it became clear that Dad was also finally slowing down professionally. He still worked at my sister's car lot but winding down to only a few hours of work a day, and then he added napping at my brother's shop to his routine. He stopped traveling to auctions or really anywhere more than an hour away. That isn't entirely surprising for an \textit{87-year-old with a 65-year career}. Everyone was happy to see him take it easy a little. Still, though, we found some of it rather strange, like all the napping. Dad hated sleeping for most of his life. He was also eating far too much sugar, even for a lifelong lover of sweets who had to have his teeth removed at 25. We heard that he was telling stories about his experience during World War II in the Italian theater, but we knew he had never served in Italy - his brother did. And through it all, he developed an even deeper distrust of banks and authority. My brother and I were informed unexpectedly and brusquely in 2018 that we could pay his bills because he was ``Done with that shit.''

By 2019, my wife and I knew there weren't too many years left. It had been another one of those long stretches where I hadn't seen Dad very often. Even though we spoke every Tuesday morning by phone, I really wanted to see him. Dad agreed to meet us for his birthday at a lovely little apple orchard - William's Orchard just outside Rural Retreat, VA - to pick apples and pumpkins with my family. He was more or less okay at this visit, albeit showing signs of being 90. It was clear he had some minor cognitive impairment: He was repeating himself a lot, he couldn't hear much at all, and he had some hygiene problems. He shared a laundry list of reasons why he didn't stray too far from home anymore and expressed his concern about how difficult it was to get to the orchard, which was only a few miles past the end of his usual range.

Something happened after that, which I mark as ``the beginning of major hostilities'' in Dad's war with cognitive decline and impairment. He had moved again to a rather remote house on the back of my brother's farm, and about six weeks after our visit to the orchard, he started acting strangely. I remember my brother calling me and saying, much to the embodiment of my fears and dread, ``Bro, you gotta come check him out. Something ain't right.'' Dad was behaving dangerously and being generally agitated, grouchy, irate, and even violent. He was always a tough man who would throw down as such when required, but now he was out of sorts every day. Paradoxically, he seemed physically fine and also ill. He threw his phone at my brother and said he wasn't going to use it anymore, which was normal, but then he really did stop using it, which was quite abnormal. He was constantly talking really, really dirty, too. He had even fired his pistol at a family member.

I hopped in my truck the next weekend I had free and paid Dad a visit. I found him in a ``We Buy Gold'' store selling collectibles that he had saved for his grandchildren. The storekeeper had been buying these for weeks off of Dad. The guy was none too happy when I reminded Dad, collected the watches, and moved us out the door.

Dad seemed completely different since I saw him at the orchard in several important ways, like selling collectibles. He believed that collectibles were actually ``things to be collected.'' He also wasn't telling stories anymore, which was devastating because he was always telling the most audacious, hilarious, thrilling, and more or less true stories from his life. These stories were like having 90 years of episodes of Dave Chappelle's ``When Keeping It Real Goes Wrong.'' He was staying very close to home, too.

It turns out that since our visit to the orchard, he had experienced two head injuries. First, he flipped his lawn mower on the hill in his yard, and a family member found him lying in the grass. (To be honest, it wasn't clear if this was before or after the orchard, but he hadn't mentioned it previously.) Second, he was working on his TV and fell into a corner, hitting his head on the wall on the way down. He had a lump on the back of his head that stuck out like an egg. He looked scared and uncertain - a new look for him - as he tried to explain what he had experienced. 

He had entered a vicious cycle of depression and loneliness where his everyday activities caused him to become distressed. For example, his loneliness made him hang out at places where he would argue with people, which just made him more depressed. The depression made him keep his distance, which made him lonely. He hadn't slept well in months because a train went by his house every night and woke him, leaving him feeling terrible and irritable every day. He was hearing voices in his house all night long, and he insisted that my sister and her kids ``were always in and out at all hours of the night partying upstairs...,'' even though they actually lived over an hour away and hadn't seen him in about a year. He described other people whom he believed to be in the house, including a beautiful young mother who was hiding from an abusive husband with her daughter and sitting in a chair next to me. (The chair was empty, but I knew who he was thinking about).  We had lots of discussions with and about people I couldn't see, various inanimate objects, and pictures. I convinced him to try out some hearing aids I brought so we could talk more, and, in turn, he tried to convince me that there were anti-Trump voices speaking whenever he flushed his toilet. To be honest, as a lifelong Democrat, I think he liked this group of auditory hallucinations the best. We had a good laugh about the toilet and went for lunch.

For the rest of the weekend and the weekends after that, I spent time with him. We had some hard discussions that fathers need to have with their children, and we laughed a lot. We buried the hatchet on a couple of issues, and he seemed to be less angry, less delusional, and to be hallucinating less. We managed to get him to a surprise 90th half birthday party in March 2020 where he got to eat ice cream with my daughter and see most of his family at once. These tough conversations and reunions were just in time, as will shortly become apparent.

I will believe until my dying days that this intervention and these visits helped him. He seemed to be  slowly getting back to how he was at the orchard. He even seemed cheerful, less depressed, and less lonely some days. He seemed happier; he had let go of some things, and the progression of his symptoms seemed to slow. He was resting more, getting a break from some of the stressors. It wasn't perfect, but it was better.

Unfortunately, my visits ended too soon as the SARS-COV-2 virus spread across the world. We all went into lockdown. I called Dad's house on Tuesday, as usual, and he asked what the hell was going on because he had just returned from the local Walmart, where he helped a woman fill her car with toilet paper. He insisted that I not visit him anymore - that I ``protect my girls.'' He knew my wife was pregnant with our second child, and he didn't want our oldest daughter to get sick, either.

In July 2020, a few months after that phone call, he had back-to-back strokes. He had walked two miles on the hottest day of the year to where he used to live to get something out of the garage. I spoke to him the next morning, and he described the strokes in detail. He was really lucid and clear during that discussion, but after that, his symptoms progressed faster than the freight train keeping him awake at night. He recovered enough to start wandering again after about six months. He was picked up in March 2021, taken to the emergency room by the county police, and soon committed to the state hospital under an Emergency Custody Order. It was almost exactly a year since I last saw him. 

It took five people to strap him to his bed when he was admitted to the hospital. He told everyone he encountered that he was going to fire them because he was the governor of the world, he was in charge, they worked for him, and he owned everything. His mind was a conflagration fueled by his outage at being hospitalized. The doctors had no idea what was going on because he didn't ``act normal for this type of patient.'' \textbf{Some of their findings over the next five months included.... Kolhman Evaluation of Living Skills....}

I appeared in court with my brother in August 2021 to petition to be guardians and conservators for Dad. A few hours later, I visited him at the care facility where he was staying. \textbf{<WALKER! Railing, etc.>} He didn't recognize me, but he thought we must be kinfolk. He was waiting on his son, who had just called and was supposed to arrive at any moment. He showed me a picture of my family so I would know who he was waiting on. When I held it up beside my face, he finally saw that I was right in front of him, although he would hardly admit to not recognizing me. 

He said it was my fault: I sure had ``got fat'' since he last saw me, and he'd recognize me better if I were skinnier! \#fml

\section*{Symptom Synopsis}

My father lived another seventeen months and continued to slowly decline over that time. The decline appeared to accelerate as he got closer to the end of his life because he lost ``big'' skills like walking, being able to feed himself, and being able to swallow solid foods. Notably, right to the last, he never forgot his name and or stopped flirting with women.

Stories or case studies like these are hard to read, but they are important because they add realism - a bit of seasoning, if you will - to more clinical descriptions. For example, I didn't really understand what a delusion looked like or how it was different than a hallucination in a real patient until doctors explained it to me with examples of Dad's behaviors. This is in spite of the fact that I had read about delusions on multiple websites and in books and heard stories about them all my life. When you read that a patient, a friend, a loved one, or whoever believed that they were "the governor of the world," it is more relatable than what Wikipedia describes as "a false fixed belief that is not amenable to change in light of conflicting evidence. 

I left the narrative above at a good place to compare to more standard descriptions because they usually don't cover what happens at the very beginning or the very end of the patient's journey. The tables below correlate my father's behaviors and symptoms with a list that I composed from other sources, including the Mayo Clinic and Wikipedia. These may look similar in others with neurocognitive disorder, or they may not present this way at all since every case is different. There are two tables: one for cognitive changes in how the brain functions (i.e., memory loss) and another for psychological changes in the way the patient behaves (i.e., inappropriate behavior). These sets of changes do not necessarily overlap (i.e., memory loss and personality changes don't necessarily happen at the same time), nor do they necessarily happen in order (i.e., personality changes don't have to precede depression).

\begin{table}[h!]
\centering
\begin{tabularx}{ 0.8\textwidth }{ | >{\raggedright\arraybackslash}X | >{\centering\arraybackslash}X | }
 \hline
 \multicolumn{2}{|c|}{Cognitive Changes} \\
 \hline
 Memory loss& No longer telling stories, forgetting that he saved collectibles for grandchildren\\
 \hline
 Problems communicating or finding words & Minimal  \\
 \hline
 Trouble with visual and spatial abilities&  Minimal \\
 \hline
 Problems with reasoning or problem-solving&  \\
 \hline
 Trouble performing complex tasks & Failed the KELS test  \\
 \hline
 Trouble with planning and organizing&   AX  \\
 \hline
 Poor coordination and control of movements & Falling repeatedly over several years  \\
 \hline
 Confusion and disorientation & Wandering, lashing out at caregivers, not recognizing family  \\
 \hline
\end{tabularx}
\caption{Table 1: Cognitive changes from the Mayo Clinic page on dementia correlated with symptoms observed in my father.}
\end{table}

\begin{table}[h!]
\centering
\begin{tabularx}{ 0.8\textwidth }{ | >{\raggedright\arraybackslash}X | >{\centering\arraybackslash}X | }
 \hline
 \multicolumn{2}{|c|}{Psychological Changes} \\
 \hline
 Personality changes& ISO ALPHA 2 Code\\
 \hline
 Depression   & AF   \\
 \hline
 Anxiety&   AX  \\
 \hline
 Agitation&   AX  \\
 \hline
 Inappropriate behavior&   AX  \\
 \hline
 Being suspicious, known as paranoia&   AX  \\
 \hline
 Hallucinating&   AX  \\
 \hline
 Delusional&   AX  \\
 \hline
\end{tabularx}
\caption{Table 2: Psychological changes from the Mayo Clinic page on dementia correlated with symptoms observed in my father.}
\end{table}

From my perspective, and I think this is supported by the narrative above, Dad experienced cognitive changes first over a timeline of about twenty years. It was straightforward and, in some cases, easy for him to conceal most of the cognitive changes if he noticed them. For example, he often waved off repeating himself or his trouble doing something as getting old, being tired, or being stressed. Those were good enough reasons for anyone who was also old, tired, or stressed. This contrasts with the seemingly sudden appearance of his psychological changes, which developed in just over a year and mostly in only a couple of months. He shared his feelings of depression and anxiety with me in October 2018. By October 2019, when we went to the orchard together, he seemed to be experiencing a few more things. However, by Thanksgiving - only a month later! - friends and family started calling me because of his behavior. When I saw him again in early January 2020, he was experiencing every psychological change in Table 2. 

Determining what is normal behavior versus psychological changes can also be extremely complicated in the moment, especially if the changes could explained by other coincident life experiences. It is tempting to retrospectively label his growing grouchiness and often over-the-line ``flirting'' from before 2018 as signs of increasing irritability, agitation, and inappropriate behaviors. That may be true, but he was also experiencing lots of normal life changes for a man in his 80s at the same time. So was he getting agitated because he was psychologically changing due to neurological disease, or was he getting pissed off that he was burning through his assets instead of working to make money because his weakening knees and aging bladder didn't let him drive as much anymore? It is extremely important to avoid jumping to conclusions that might impact someone's quality of life or remove their autonomy and personal freedoms without first examining their situation. And, of course, on the other hand, there is a point where it becomes apparent that there are deeper problems - as it did for us in 2019 - when the changes in behavior are so absurd that the only rational conclusion can be that something is wrong.

What to do once it is clear that something is wrong is where we will pick up the story next time.

\section*{Next Time}

Thanks for reading this second part of a series about my father’s struggles with cognitive decline and what we did to help him. I would especially like to thank my wonderful wife Paige and my dear friend Mike Phelan for encouraging me to finish this article and for those who reached out after the first article with kind words of encouragement or thanks. I hope you’ll come back to read the next part, in which I'll discuss ways families can get help to support their loved ones, including public and private organizations, social support services, legal support for elder care, how and when to engage law enforcement, and how to have tough conversations with family members.

%First, no two patients with dementia are the same. Each patient takes a different path and develops a unique expression of symptoms. Second, whether it starts abruptly due to head trauma or is a slow burn, dementia evolves over time. Third, some degree of mild cognitive impairment might be easy to accommodate. Fourth, and this is the important one, you can help your friend or family member at any and all stages of dementia. They are just as lovable on the last day as they were on day 1. My Dad was still ``Jack'' right until the end.

%Basically: Here's how it looked behind the scenes, here's how he masked it, and here's what we wished we would have known about public support. Including that social services isn't out to get you.

%No stories, physical symptoms, depression, strong delusions, strong hallucinations, violent behavior, extremely dirty language, poor decision making, bad hygiene, trouble counting anything but money.

%oss of business acumen, ability to plan, trouble getting home. Problems with authority. Integrity issues.

%Repetitive speech, stronger delusions, personal hygiene.

\end{document}
